Quantifying the ease of viral escape from broad and narrow antibodies to influenza hemagglutinin

### 2017-10-22 paper/manuscript.tex
a few minor revisions to results

\subsection*{Quantifying the antigenic effects of all mutations selected by each antibody}
We performed mutational antigenic profiling using the three broad antibodies at the concentrations indicated in Figure~\ref{fig:neutcurves}, and calculated the fraction of virions carrying each mutation that survived neutralization. 
All experiments were performed in full biological triplicate using three independently generated virus libraries carrying all single amino-acid mutations to HA~\citep{doud2016accurate}.
%We have previously described~\citep{doud2017complete} mutational antigenic profiling of the three narrow strain-specific antibodies, and we re-analyzed those deep sequencing data to calculate the fraction of virions with each mutation that survived neutralization.
The correlations among replicates are in Figure~\ref{suppfig:corr}.
For the remainder of this paper, we will refer to the median antigenic effect of each mutation across replicates.

@@

For the broad anti-receptor binding pocket antibody S139/1, we observe strong escape mutations at sites 156, 158, and 193 (Figure~\ref{fig:structures}A; sites are in H3 numbering). 
These three sites fall directly in the physical binding footprint of the antibody~\citep{lee2012heterosubtypic}, and are the same three sites where \citet{yoshida2009cross} selected escape mutants in H1, H2, and H3 HAs. 
Our data show that numerous different amino-acid mutations at each site confer neutralization resistance.
The mutation with the largest effect, D158N, introduces an N-linked glycosylation motif.

@@

However, C179 selects several mutations that do not have any apparent effect on FI6v3 (Figure~\ref{fig:structures}A, Figure~\ref{suppfig:FI6v3logo}).
The most notable of these C179-specific mutations occur at site 38.
The additional breadth of FI6v3 over antibodies such as C179 that neutralize only group 1 HAs is because FI6v3 can accommodate a glycan on the asparagine at site 38 that is present in group 2 HAs~\citep{corti2011neutralizing,sui2009structural,ekiert2009antibody}. 
However, the H38S mutation that has the largest effect on C179 resistance in our experiments does not introduce a glycosylation motif, showing that there are also other ways to escape anti-stalk antibodies at this site.
Interestingly, group 1 HA subtypes that are susceptible by C179 tend to possess a histidine at site 38, but subtypes that are not bound or neutralized by C179 often possess a serine (Figure~\ref{fig:structures}C). 
%The amino-acid identity at site 111 of HA2 can change the orientation of a conserved Trp21 in HA2, also resulting in group 1 and group 2 differences in the binding ability of stalk-targeting antibodies. An H111T mutation in the A/South Carolina/1/1918 (H1N1) has been reported to abrogate C179 binding~\citep{dreyfus2013structure}.
%However, we did not observe escape mutations at site 111, but this may be due to strain-specific differences.

@@

\begin{figure}
\centerline{\includegraphics[width=0.8\textwidth]{figs/FI6v3mutant_neutcurves/FI6v3_mutant_neutcurves.pdf}}
\caption{
\label{fig:FI6v3neutcurves}
{\bf The mutations selected by FI6v3 increase neutralization resistance, but the effects are small.}
(A) Neutralization curves of individual viral mutants with FI6v3.
The mutations K280S, K280T, N291S, G47R (HA2), and K(-8)T are all expected to increase neutralization resistance based on the mutational antigenic profiling (Figure~\ref{fig:structures}A), and they all do so -- but only modestly. 
The mutations K280A, M17L (HA2), P80D, and V135T are \emph{not} expected to affect neutralization based on the mutational antigenic profiling (Figure~\ref{suppfig:FI6v3logo}, and none of them do.
(B), (C) In contrast to FI6v3, mutations selected by other antibodies have very large effects on neutralization.
Shown are neutralization curves for representative escape mutants from H17-L19 and H17-L7 taken from \citet{doud2017complete}; that paper also shows additional examples of large-effect escape mutations from these antibodies.
}
\end{figure}

Figure~\ref{fig:FI6v3neutcurves} shows that the mutational antigenic profiling is highly predictive of the results of the neutralization assays, even for small-effect mutations.
As discussed in the previous section, the three mutations most strongly selected by FI6v3 introduce glycosylation motifs at sites 278-280 or 289-291 (Figure~\ref{fig:structures}A).
We created viruses carrying each of these mutations (K280S, K280T, and N291S) and validated that all three mutants were modestly more resistant to FI6v3 (Figure~\ref{fig:FI6v3neutcurves}A).
As a control, we also validated that a mutation at one of these sites (K280A) that does \emph{not} have an effect in our mutational antigenic profiling does not shift the neutralization curve (Figure~\ref{fig:FI6v3neutcurves}A).
Interestingly, the HA of pandemic H1N1 has a glycosylation motif at 278-280, and \citet{corti2011neutralizing} have reported that this HA is $\sim$4-fold more resistant to FI6v3 than the A/WSN/1933 HA used in our experiments.
This reported difference between strains is of similar magnitude to the effect that we measure for introducing the glycosylation motif via the K280S or K280T mutations.

Our mutational antigenic profiling also identified several non-glycosylation-motif mutations that were selected by FI6v3.
We validated that one of these mutations, G47R in the HA2 chain, indeed increased neutralization resistance (Figure~\ref{fig:FI6v3neutcurves}A) -- although as predicted by the mutational antigenic profiling, the magnitude of the effect was smaller than for the glycosylation-motif mutations.
The most unexpected mutations identified in the mutational antigenic profiling were at site -8 in the signal peptide.
We tested one of these mutations, K(-8)T, and found that it indeed caused a modest but detectable increase in neutralization resistance (Figure~\ref{fig:FI6v3neutcurves}A).
As controls, we also tested three more mutations (P80D, V135T, and M17L in HA2) that did \emph{not} have substantial effects in the mutational antigenic profiling, and confirmed that none of them affected neutralization resistance.

A remarkable aspect of these validation experiments is the very small effect sizes of the identified mutations.
Antigenic mutations selected by strain-specific antibodies to HA generally increase the concentration of antibody needed to neutralize the virus by orders of magnitude.
Neutralization curves for a few examples of such large-effect escape mutants are in Figure~\ref{fig:FI6v3neutcurves}B,C.
Our work demonstrates that there are no such large-effect single mutations to the A/WSN/1933 HA when it comes to escaping from FI6v3 and C179.
But the results in Figure~\ref{fig:FI6v3neutcurves}A show that we can still leverage the high sensitivity of mutational antigenic profiling to identify mutations that have much more modest but nonetheless measurable effects on antibody neutralization.

>>>>

\subsection*{Quantifying the antigenic effects of all mutations selected by each antibody}
We performed mutational antigenic profiling using the three broad antibodies at the concentrations indicated in Figure~\ref{fig:neutcurves}, and calculated the fraction of virions carrying each mutation that survived neutralization. 
All experiments were performed in full biological triplicate using three independently generated virus libraries carrying single amino-acid mutations to HA~\citep{doud2016accurate}.
%We have previously described~\citep{doud2017complete} mutational antigenic profiling of the three narrow strain-specific antibodies, and we re-analyzed those deep sequencing data to calculate the fraction of virions with each mutation that survived neutralization.
The correlations among replicates are in Figure~\ref{suppfig:corr}.
For the remainder of this paper, we will refer to the median antigenic effect of each mutation across replicates.

@@

For the broad anti-receptor binding pocket antibody S139/1, there are strong escape mutations at sites 156, 158, and 193 (Figure~\ref{fig:structures}A; sites are in H3 numbering). 
These three sites fall directly in the physical binding footprint of the antibody~\citep{lee2012heterosubtypic}, and are the same three sites where \citet{yoshida2009cross} selected escape mutants in H1, H2, and H3 HAs. 
Our data show that numerous different amino-acid mutations at each site confer neutralization resistance.
The mutation with the largest effect, D158N, introduces an N-linked glycosylation motif.

@@

However, C179 selects several mutations that do not have any apparent effect on FI6v3 (Figure~\ref{fig:structures}A, Figure~\ref{suppfig:FI6v3logo}).
The most notable of these C179-specific mutations are at site 38.
The additional breadth of FI6v3 over antibodies such as C179 that neutralize only group 1 HAs is because FI6v3 can accommodate a glycan on the asparagine at site 38 that is present in group 2 HAs~\citep{corti2011neutralizing,sui2009structural,ekiert2009antibody}. 
However, the H38S mutation that has the largest effect on C179 resistance in our experiments does not introduce a glycosylation motif, showing that there are also other ways to escape anti-stalk antibodies at this site.
Interestingly, group 1 HA subtypes that are susceptible to C179 tend to possess a histidine at site 38, but subtypes that are not bound or neutralized by C179 often possess a serine (Figure~\ref{fig:structures}C). 
%The amino-acid identity at site 111 of HA2 can change the orientation of a conserved Trp21 in HA2, also resulting in group 1 and group 2 differences in the binding ability of stalk-targeting antibodies. An H111T mutation in the A/South Carolina/1/1918 (H1N1) has been reported to abrogate C179 binding~\citep{dreyfus2013structure}.
%However, we did not observe escape mutations at site 111, but this may be due to strain-specific differences.

@@

\begin{figure}
\centerline{\includegraphics[width=0.8\textwidth]{figs/FI6v3mutant_neutcurves/FI6v3_mutant_neutcurves.pdf}}
\caption{
\label{fig:FI6v3neutcurves}
{\bf The mutations selected by FI6v3 increase neutralization resistance, but the effects are small.}
(A) Neutralization curves of individual viral mutants with FI6v3.
The mutations K280S, K280T, N291S, G47R (HA2), and K(-8)T are all expected to increase neutralization resistance based on the mutational antigenic profiling (Figure~\ref{fig:structures}A), and they all do so -- but only modestly. 
The mutations K280A, M17L (HA2), P80D, and V135T are \emph{not} expected to affect neutralization based on the mutational antigenic profiling (Figure~\ref{suppfig:FI6v3logo}), and none of them do.
(B), (C) In contrast to FI6v3, mutations selected by other antibodies have very large effects on neutralization.
Shown are neutralization curves for representative escape mutants from H17-L19 and H17-L7 taken from \citet{doud2017complete}.
}
\end{figure}

Figure~\ref{fig:FI6v3neutcurves} shows that the mutational antigenic profiling is highly predictive of the results of the neutralization assays, even for small-effect mutations.
As discussed in the previous section, the three mutations most strongly selected by FI6v3 introduce glycosylation motifs at sites 278-280 or 289-291 (Figure~\ref{fig:structures}A).
We created viruses carrying each of these mutations (K280S, K280T, and N291S) and validated that all three modestly increased resistance to FI6v3 (Figure~\ref{fig:FI6v3neutcurves}A).
As a control, we also validated that a mutation at one of these sites (K280A) that does \emph{not} have an effect in our mutational antigenic profiling does not shift the neutralization curve (Figure~\ref{fig:FI6v3neutcurves}A).
Interestingly, the HA of pandemic H1N1 has a glycosylation motif at 278-280, and \citet{corti2011neutralizing} have reported that this HA is $\sim$4-fold more resistant to FI6v3 than the A/WSN/1933 HA used in our experiments.
This reported difference between strains is of similar magnitude to the effect that we measure for introducing the glycosylation motif via the K280S or K280T mutations.

Our mutational antigenic profiling also identified several non-glycosylation-motif mutations that were selected by FI6v3.
We validated that one of these mutations, G47R in the HA2 chain, increased neutralization resistance (Figure~\ref{fig:FI6v3neutcurves}A) -- although as predicted by the mutational antigenic profiling, the magnitude of the effect was small.
The most unexpected mutations identified in our mutational antigenic profiling were at site -8 in the signal peptide.
We tested one of these mutations, K(-8)T, and found that it indeed caused a modest but detectable increase in neutralization resistance (Figure~\ref{fig:FI6v3neutcurves}A).
As controls, we also tested three more mutations (P80D, V135T, and M17L in HA2) that did \emph{not} have substantial effects in the mutational antigenic profiling, and confirmed that none of them affected neutralization resistance.

A notable aspect of these validation experiments is the very small effect sizes of the identified mutations.
Antigenic mutations selected by strain-specific antibodies to HA generally increase the concentration of antibody needed to neutralize the virus by orders of magnitude.
Neutralization curves for examples of such large-effect escape mutants are in Figure~\ref{fig:FI6v3neutcurves}B,C.
Although there are no such large-effect single mutations that escape FI6v3 and C179, the results in Figure~\ref{fig:FI6v3neutcurves}A show that we can still use mutational antigenic profiling to identify mutations that have small but measurable effects on resistance to these antibodies.
